Automatic Matching of Bullet Lands

### 2016-07-25 revision/response.Rmd

---
title: "Response to reviewers"
author: "Heike Hofmann, Eric Hare, Alicia Carriquiry"
date: "July 26, 2016"
output: html_document
---

@@

***We selected random forests because the method has desirable attributes and can be easily ported to other forensic applications. We were not intending to provide new insights on random forests. The focus of the paper was to show the potential of combining a large number ofimage attributes to discriminate among bullets. As the forest ultimately predicted every land-to-land comparison properly, we felt that a comparison to other learning methods was outside the scope of this paper. Had a wider variety of data been available, there would likely be more variability in the predictions and a comparison to other learning methods would be prudent.***

# Reviewer B

4. The authors should discuss their work in the context of Nicholas Petraco’s work as well as other ongoing research endeavors in impression and pattern evidence. See https://www.ncjrs.gov/pdffiles1/nij/grants/239048.pdf The alignment method should have been compared with a standard method- See Patraco’s NIJ report for a couple of examples. As presented the novel component of the paper is the "twicing" for grooves combined with a loess fit to the remove curves from bullet images. I am not sure this is the authors intention. It would be interesting to compare this method with some of the classical methods from functional data analysis used for curve registration.

***Using the cross correlation function as a method of alignment is actually pretty standard - the appropriate references are now included.***

***chumbley:2010 uses 1d profiles and discusses several different angles of extraction (no alignment necessary, no grooves and no flattening) chubley:2010 is a much more detailed version over David Faden's thesis***

@@

15. On a related subject, I believe that a random forest provides the probability that two bullets were fired by the same gun given the set of observations (Fig 19), not the probability of interest: the probability of two sets of observations given that the two bullets were fired by the same gun. The authors may want to revise some of the statements in their discussion and conclusion.

***The reviewer is correct in stating that two probabilities of interest are the probability of observing the data under the "same source" (same gun) hypothesis or under the "different source" hypothesis.  An empirical approach to get at those two probabilities is to derive the distribution of scores when two bullets are known to have been fired by the same gun and the distribution of scores when two bullets are known to have been fired by different guns.  We have actually derived those to distributions and include them in the revised manuscript.***

16. I am concerned by the assumption of independence of between the different lands on a given bullet that the authors seems to use in their error rates calculations (and ROC curves). The assumption seems unreasonable since as soon as a land has been paired between two bullets, it seems that it becomes more likely to pair the other lands. In other words, I believe that the authors need to redo their error rate measurements, but using bullets as their experimental units and not lands. I am also concerned by the overwhelming majority of non-matches, this must certainly artificially increase the rate of correct exclusion and distort the data.

***Assuming independence between the lands of each barrel is certainly not realistic in practice, however, by assuming independence we make the task of matching harder for the algorithm. This also has the benefit of potentially being more useful for degraded data, where we may not have a full set of six groove-to-groove scans. Based on the actual numerical values of features derived from each of the lands, an independence assumption is not unrealistic, because the features do not exhibit any clustering associated to barrel (Bachrach 2006). The technique suggested by the reviewer could be solved using a hierarchical approach to matching, and we agree with the assessment that it would likely decrease the error rates.***

### 2016-07-25 imaging-paper.Rnw
More responses

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lof$bulletname <- scrubPath(lof$bullet)
qplot(y, resid, group=bulletname, colour=bulletname, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lofY$bulletname <- scrubPath(lofY$bullet)
qplot(y+min(lof$y), resid, group=bulletname, colour=bulletname, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
\hh{For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures~\cite{bachrach:2002, vorburger:2011, thompson:2013}.}
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

>>>>

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available. Modern automated techniques using 3D images have also been conducted by \citet{riva:2014}, but this work applies to striae found on cartridge cases and not bullets.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lof$bulletname <- scrubPath(lof$bullet)
qplot(y, resid, group=bulletname, colour=bulletname, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lofY$bulletname <- scrubPath(lofY$bullet)
qplot(y+min(lof$y), resid, group=bulletname, colour=bulletname, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
\hh{For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures~\cite{bachrach:2002, chu:2010, vorburger:2011, thompson:2013}.}
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

### 2016-07-25 imaging-paper.tex
More responses

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. \eh{We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging. XXX need more?}
The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate, no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available.

@@

\end{knitrout}
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
\hh{For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures~\cite{bachrach:2002, vorburger:2011, thompson:2013}.}
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

@@

\end{knitrout}
\caption{\label{fig:varysmooth} Peak/valley detection at smoothing factors of 5, 25, and 45, respectively. Note that a smoothing factor of 5 yields enough noise that many very minimal overlapping peaks and valleys are detected, while a smoothing factor of 45 might over-smooth and cause the peaks/valleys to either end disappear or shift  horizontally  from their position in the signature.}
\end{figure}

@@

Using recursive partitioning, we fit a decision tree~\citep{breiman:1984, rpart, rpart.plot} to predict matches between lands based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of 132 lands is being matched correctly. Interestingly, the number of consecutive matching striae does not feature in this evaluation. 
Instead of CMS, cross-correlation (ccf) between the signatures is very important in the matching process  by the decision tree. Aside from  cross correlation, the total number of matches is also included in the decision rule. 
Between cross-correlation and CMS, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2D microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitized version of the images that preserves the 3D surface structure. The findings about the discriminating power of cross-correlation is consistent with the results of the study by \citet{ma:2004}. However, in that study, authors did not consider the number of matches and non-matches.

>>>>

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. Our algorithm is fully open source and available on GitHub \citep{x3prplus}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, examine the factors that separate known bullet matches from non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.
The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate, no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available. Modern automated techniques using 3D images have also been conducted by \citet{riva:2014}, but this work applies to striae found on cartridge cases and not bullets.

@@

\end{knitrout}
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
\hh{For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures~\cite{bachrach:2002, chu:2010, vorburger:2011, thompson:2013}.}
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

@@

\end{knitrout}
\caption{\label{fig:varysmooth} Peak/valley detection at smoothing factors of 5, 25, and 45, respectively. Note that a smoothing factor of 5 yields enough noise that many very minimal overlapping peaks and valleys are detected, while a smoothing factor of 45 might over-smooth and cause the peaks/valleys to either end disappear or shift  horizontally from their original position in the signature.}
\end{figure}

@@

Using recursive partitioning, we fit a decision tree~\citep{breiman:1984, rpart, rpart.plot} to predict matches between lands based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of 132 lands is being matched correctly. Interestingly, the number of consecutive matching striae does not feature in this evaluation. 
Instead of CMS, cross-correlation (ccf) between the signatures is very important in the matching process  by the decision tree. Aside from  cross correlation, the total number of matches is also included in the decision rule. 
Between cross-correlation and CMS, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2D microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitized version of the images that preserves the 3D surface structure. The findings about the discriminating power of cross-correlation are consistent with the results of the study by \citet{ma:2004}. However, in that study, the authors did not consider the number of matches and non-matches.
